\Huge{Vipassana for Hackers} \\ \large{Paper Two: The Brain} \\ \large\textit{Version 0.2}

### 2019-09-30 paper-two/paper-two.tex
added information about limbic system and ssc.with references

The DMN is comprised of specific anatomy including portions of the mid-line of the
brain, an evolutionarily primitive area related to memory and emotion, and
structures in the cortex, an evolutionarily recent part of the brain containing the
executive and higher order functions. \cite{defaultnetworkanatomy} These functionally
connected regions are involved in the neurological basis
of the self, considering the mental states of others, remembering the past, and
imagining the future. 
This network is mostly observed through changes in blood flow to different parts of 
the brain measured using fMRI (functional Magnetic Resonance Imaging) or 
PET (Positron Emission Tomography){defaultnetworkadaptive}

@@

The \textit{Theory of Mind} is a body of work in philosophy and psychology which
describes an individual's ability to consider the mental states of others, such as
appreciating another individual's false beliefs (knowledge based on incorrect or
outdated information). The Theory of Mind captures not only one's ability to
attribute beliefs, desires, and emotions to others but also to oneself --- and to
differentiate the two. \cite{autistictheoryofmind}
There is a rich body of work pointing to this network (especially the dmPFC) being engaged
when individuals reflect upon, evaluate, appraise their own thoughts, feelings and desires 
as well as those of others. \cite{theoryofmind}

@@

@@

So, the Central Executive Network is involved in goal-directed behavior and processes 
related to attention. These goals are usually external and attention is also directed externally(whether 
voluntary or not). When you meditate, you direct your attention voluntarily
inside, to the same sensory cues that are directing the PPC.
When you meditate, you also cultivate your cognitive processes related to attention.
When your wanders, you notice it and bring it back.

\todo{Summarize CEN? Relevance to meditation?}

@@

The AIC and ACC together give rise to our interoceptive and conscious
self-awareness. \todo{add reference justifying statement that AIC+ACC create
  conscious self-awareness}

\subsection{The Limbic System}

The limbic system is a complex set of structures that lies on both sides of the
Thalamus, just under the Cerebrum. It includes the Hypothalamus, the Hippocampus, the
Amygdala, and several other nearby areas. It appears to be primarily responsible for
our emotional life, and has a lot to do with the formation of memories. \todo{add
  reference for limbic system}

@@

And so it was. With my brain filled with thoughts about myself, about the people
around me, I started this meditation business. For the first 3.5 days you are asked
to focus your attention on the area below your nostrils and above your upper
lip. That's all. Within the first two days, I had made up elaborate stories about my
fellow meditators, all of them superheroes, tirelessly working to save humanity
together. My default mode network was in overdrive.

@@

(The somatosensory cortex is located in the gyrus and the parietal lobes: The sensory
homunculus which is a part of the somatosensory cortex is a cortical representation
of the body based on the degree of sensory innervation) \todo{Maybe move sensory
  homunculus diagram and explanation to the anatomy section?}

@@

@@

@@

>>>>

The DMN is comprised of specific anatomy including portions of the mid-line of the
brain, an evolutionarily primitive area related to memory and emotion, and
structures in the cortex, an evolutionarily recent part of the brain containing the
executive and higher order functions. \cite{defaultnetworkanatomy}

These functionally connected regions are involved in the neurological basis
of the self, considering the mental states of others, remembering the past, and
imagining the future.

This network is mostly observed through changes in blood flow to different parts of 
the brain measured using fMRI (functional Magnetic Resonance Imaging) or 
PET (Positron Emission Tomography){defaultnetworkadaptive}

@@

The \textit{Theory of Mind} is a body of work in philosophy and psychology which
describes an individual's ability to consider the mental states of others, such as
appreciating another individual's false beliefs (knowledge based on incorrect or
outdated information). The Theory of Mind captures not only one's ability to
attribute beliefs, desires, and emotions to others but also to oneself --- and to
differentiate the two. \cite{autistictheoryofmind}

There is a rich body of work pointing to this network (especially the dmPFC subsystem) 
being engaged when individuals reflect upon, evaluate, appraise their own thoughts, 
feelings and desires as well as those of others. \cite{theoryofmind}

@@

The Medial Temporal Subsytem is primarily activated during these processes
This subsystem consists of the Medial Temporal Lobe and parts of the limbic system
especially the Hippocampus.

The limbic system is a complex set of structures that lies on both sides of the
Thalamus, just under the Cerebrum. It includes the Hypothalamus, the Hippocampus, the
Amygdala, and several other nearby areas. It appears to be primarily responsible for
our emotional life, and has a lot to do with the formation of memories. \cite{limbic}

@@

\subsection {Sidenote: Somatosensory Cortex}
The primary somatosensory cortex is located in a ridge of cortex found within the parietal lobe.
It is responsible for processing somatic sensations. These sensations arise from receptors 
positioned throughout the body. They are responsible for detecting touch, for proprioception 
(i.e. the position of the body in space), nociception (i.e. pain), and temperature. \cite{somato}
When such receptors detect one of these sensations,  the information is sent to the thalamus and 
then to the primary somatosensory cortex.
The sensory homunculus which is a part of the somatosensory cortex is a cortical representation
of the body based on the degree of sensory innervation. \cite{sensoryhom}

\todo {include images of the homunculus here}

\todo{Summarize CEN? Relevance to meditation?} 
So, the Central Executive Network is involved in goal-directed behavior and processes 
related to attention. These goals are usually external and attention is also directed externally(whether 
voluntary or not). When you meditate, you direct your attention voluntarily
inside, to the same sensory cues that are directing the PPC.

@@

The AIC and ACC together give rise to our interoceptive and conscious
self-awareness.  \cite{selfaware}

The salience network is directly involved in switching between the DMN and CN. 
When you meditate, you cultivate the cognitive processes related to attention.
When your mind wanders, you notice it and bring it back. 
The salience network should be directly involved in this process of assigning attention. 
Meditation should strengthen the salience network and its ability to switch from
the default.

@@

And so it was. With my brain filled with thoughts about myself, about the people
around me, I started this meditation business. For the first 3.5 days you are asked
to focus your attention on the area below your nostrils and above your upper
lip. That's all, these were really simple instructions, but so hard to keep up with.
Within the first two days, I had made up elaborate stories about my
fellow meditators, all of them superheroes, tirelessly working to save humanity
together. My default mode network was in overdrive.

@@

(The somatosensory cortex is located in the gyrus and the parietal lobes:) \todo{Maybe move sensory
  homunculus diagram and explanation to the anatomy section?}

@@

\bibitem{limbic}
  Catani M, Dell'Acqua F, Thiebaut De Schotten M
  \textit {A revised limbic system model for memory, emotion and behaviour}
  Neuroscience and Biobehavioral Reviews. 37 (8): 1724–37. 
  doi:10.1016/j.neubiorev.2013.07.001. PMID 23850593.

@@

\bibitem{selfaware}
  Anil K. Seth, Keisuke Suzuki, Hugo D. Critchley
  \textit {An Interoceptive Predictive Coding Model of Conscious Presence}
  Front Psychol. 2011; 2: 395.

@@

\bibitem{somato}
  Viaene A.N. 
  \textit {synaptic properties of thalamic input to layers 2/3 and 4 of 
              primary somatosensory and auditory cortices}
  Journal of Neurophysiology. 105 (1): 279–292. 
  doi:10.1152/jn.00747.2010. PMC 3023380.

\bibitem{sensoryhom}
  Marieb E, Hoehn K. 
  Human Anatomy and Physiology. 
  7th Ed. 2007. Pearson Benjamin Cummings: San Francisco.
